---
title: "PS5: INFO 201"
author: "Kaden Kapadia"
date: "2/13/2023"
output: html_document
---

# 1 Load and check data (5pt)
You first task is to do a very simple data check: 

1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See [**https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating**](https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating). \
- title: "PS5: INFO 201"
- author: "Kaden Kapadia"
- date: "2/13/2023"

2. (2pt) Load data. How many rows/columns do we have?  \
```{r, message = FALSE}
library(tidyverse)
gapminder <- read_delim("gapminder.csv.bz2")
rows <- gapminder %>% 
  nrow()
cols <- gapminder %>% 
  ncol()
cat("There are", rows, "rows and", cols, "columns in the data.")
```

3. (2pt) Print a small sample of data. Does it look OK? 
```{r}
head(gapminder)
# Yes, the data looks good.
```
---------------------------------------------------------------------


# 2 Descriptive statistics (15pt)

1. (3pt) How many countries are there in the dataset? Analyze all three: *iso3*, *iso2* and *name*.
```{r}
gapminder %>% 
  select(iso3, iso2, name) %>% 
  summarise(iso3 = n_distinct(iso3), iso2 = n_distinct(iso2), name = n_distinct(name))
```

2. If you did this correctly, you saw that there are more iso-2 codes than names, and there are even more *iso3*-codes. What is going on? Can you find it out?

    (a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?
```{r}
gapminder %>% 
  group_by(iso2) %>% 
  summarise(num_name = n_distinct(name)) %>% 
  arrange(desc(num_name))
```

    (b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries? Hint: two of these entitites are *CHANISL* and *NLD CURACAO*.
```{r}
gapminder %>% 
  group_by(name) %>% 
  summarise(num_iso3 = n_distinct(iso3)) %>% 
  arrange(desc(num_iso3))
```
  
3. (2pt) What is the minimum and maximum year in these data?
```{r}
min_year <- min(as.numeric(gapminder$time), na.rm = TRUE)
max_year <- max(as.numeric(gapminder$time), na.rm = TRUE)
cat("Minimum year:", min_year, "\n")
cat("Maximum year:", max_year, "\n")
```
  
---------------------------------------------------------------------

# 3 CO2 emissions (30pt)
Next, let’s analyze CO2 emissions.

1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing *CO2* and *co2_PC*. Which years have most missing data? \
```{r}
gapminder %>%
  group_by(time) %>%
  summarise(num_missing_co2 = sum(is.na(co2)), num_missing_co2_pc = sum(is.na(co2_PC))) %>%
  arrange(desc(num_missing_co2), desc(num_missing_co2_pc))
```

2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see. \
```{r}
countries <- c("United States of America", "China", "India", "Uruguay", "Sudan")
co2_data <- gapminder %>%
  filter(name %in% countries)
ggplot(data = co2_data, aes(x = time, y = co2, color = name)) +
  geom_line() +
  labs(x = "Year", y = "Total CO2 Emissions", color = "Country") +
  theme_bw()
```

3. (5pt) Now let’s analyze the CO2 emissions per capita (*co2_PC*). Make a similar plot of thesame countries. What does this figure suggest? \
```{r}
ggplot(data = co2_data, aes(x = time, y = co2_PC, color = name)) +
  geom_line() +
  labs(x = "Year", y = "Total CO2 Emissions Per Capita", color = "Country") +
  theme_bw()
```

4. (6pt) Compute average CO2 emissions per capita across the continents (assume *region* is the same as continent). Comment what do you see. \
Note: just compute averages over countries and ignore the fact that countries are of different size. \
Hint: Americas 2016 should be 4.80. \
```{r}
gapminder %>%
  filter(time == 2016) %>% 
  group_by(region) %>%
  summarise(avg_co2_PC = mean(co2_PC, na.rm = TRUE))
```

5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016. \
Hint: it should look something along these lines(**Please refer to the spec for the graph**) \
```{r}
co2_continent <- gapminder %>% 
  filter(!is.na(region)) %>% 
  filter(time %in% c(1960, 2016)) %>% 
  group_by(time, region) %>% 
  summarise(avg_co2_PC = mean(co2_PC, na.rm = TRUE))
ggplot(co2_continent, aes(x = region, y = avg_co2_PC, fill = as.factor(time))) + 
  geom_bar(stat = "identity", position = "dodge") + 
  labs(x = "Continent", y = "Average CO2 Emissions per Capita", 
       title = "Average CO2 Emissions per Capita Across Continents in 1960 and 2016",
       fill = "Year")
```

6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume *region* is continent). \
```{r}
gapminder %>%
  filter(!is.na(co2_PC)) %>% 
  filter(time == 2016) %>%
  group_by(region) %>%
  arrange(co2_PC) %>%
  slice_head(n = 3) %>%
  select(region, name, co2_PC) %>%
  bind_rows(
    gapminder %>%
      filter(time == 2016) %>%
      group_by(region) %>%
      arrange(desc(co2_PC)) %>%
      slice_head(n = 3) %>%
      select(region, name, co2_PC)
  ) %>%
  arrange(region, co2_PC)
# Because 2019 data had many NA values, I chose to use 2016 data instead.
```
---------------------------------------------------------------------

# 4 GDP per capita (50pt)
Let’s look at GDP per capita (*GDP_PC*)

1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better. \
Comment what do you see there. \
```{r}
filter_gapminder <- gapminder %>% 
  filter(!is.na(GDP_PC))
ggplot(data = subset(filter_gapminder, time == 1960),
       aes(x = GDP_PC, y = lifeExpectancy)) +
  geom_point(aes(size = totalPopulation, color = region)) +
  labs(x = "GDP per capita", y = "Life expectancy", title = "GDP per capita versus 
        life expectancy by country in 1960")
```

2. (4pt) Make a similar plot, but this time use 2019 data only. \
```{r}
filter_gapminder <- gapminder %>% 
  filter(!is.na(GDP_PC))
ggplot(data = subset(filter_gapminder, time == 2019),
       aes(x = GDP_PC, y = lifeExpectancy)) +
  geom_point(aes(size = totalPopulation, color = region)) +
  labs(x = "GDP per capita", y = "Life expectancy", title = "GDP per capita versus 
        life expectancy by country in 2019")
```

3. (6pt) Compare these two plots and comment what do you see. How has world developed through the last 60 years? \
```{r}
```

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures? \
Note: here as *average*I mean just average over countries, ignore the fact that countries are of different size.\
```{r}
avg_le_1960 <- gapminder %>% 
  filter(!is.na(lifeExpectancy)) %>% 
  filter(time == 1960) %>% 
  group_by(region) %>% 
  summarise(avg_le_1960 = mean(lifeExpectancy))
avg_le_1960

avg_le_2019 <- gapminder %>% 
  filter(!is.na(lifeExpectancy)) %>% 
  filter(time == 2019) %>% 
  group_by(region) %>% 
  summarise(avg_le_2019 = mean(lifeExpectancy))
avg_le_2019
```

5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see. \
Hint: these data (data in long form) is not the simplest to compute growth. But you may want to check out the lag() function. And do not forget to group data by continent when using lag(), otherwise your results will be messed up! See [**https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute**](https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute).\
```{r}
gapminder %>%
  filter(!is.na(lifeExpectancy), time >= 1960 & time <= 2019) %>% 
  filter (!is.na(name)) %>% 
  group_by(region, name) %>% 
  summarise(LE_growth = lifeExpectancy - lag(lifeExpectancy)) %>% 
  group_by(region) %>% 
  summarise(avg_LE_growth = mean(LE_growth, na.rm = TRUE)) %>% 
  arrange(avg_LE_growth)
```

6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it! \
```{r}
gapminder %>% 
  filter(time == 1960 | time == 2019, !is.na(GDP_PC)) %>% 
  ggplot(aes(x = GDP_PC, fill = as.factor(time))) + 
  geom_histogram(alpha = 1, position = "stack") + 
  labs(x = "GDP per capita", y = "Frequency Distribution",
       title = "GDP per capita for 1960 and 2019", fill = "Year")
```

7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.) \
Hint: check out the function rank()! \
Hint2: 17 for 1960. \
```{r}
gapminder %>% 
  filter (time == 1960, !is.na(lifeExpectancy)) %>% 
  arrange (desc(lifeExpectancy) ) %>% 
  mutate (ranking_1960 = rank(desc(lifeExpectancy) ) ) %>% 
  filter (name == "United States of America") %>% 
  summarise (name, ranking_1960)
gapminder %>% 
  filter (time == 2019, !is.na(lifeExpectancy)) %>% 
  arrange (desc(lifeExpectancy) ) %>% 
  mutate (ranking_2019 = rank(desc(lifeExpectancy) ) ) %>% 
  filter (name == "United States of America") %>% 
  summarise (name, ranking_2019)
```

8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year? \
Hint: 0.0904 for 1960. \
```{r}
countries_1960 <- gapminder %>% 
  filter(time == 1960, !is.na(lifeExpectancy)) %>% 
  filter(name != "") %>% 
  summarise(num_countries_1960 = n())
  
countries_2019 <- gapminder %>% 
  filter(time == 2019, !is.na(lifeExpectancy)) %>% 
  filter(name != "") %>% 
  summarise(num_countries_2019 = n())
  
relative_rank_1960 <- gapminder %>% 
  filter(time == 1960) %>% 
  filter(name != "") %>% 
  mutate (rank_1960 = rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America") %>% 
  pull(rank_1960)/countries_1960$num_countries_1960
  
relative_rank_2019 <- gapminder %>% 
  filter(time == 2019) %>% 
  filter(name != "") %>% 
  mutate (rank_2019 = rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America") %>% 
  pull(rank_2019)/countries_2019$num_countries_2019
  
cat("The relative rank of the United States in 1960 is:", relative_rank_1960, "\n")
cat("The relative rank of the United States in 2019 is:", relative_rank_2019, "\n")
```
---------------------------------------------------------------------
Finally tell us how many hours did you spend on this PS.
I spent 7 hours on this problem set. 